Hierarchical Principal Manifold Estimation

### 2023-08-11 paper/main.tex
add model specification

With this background in place, we turn to the motivating situation of analyzing the subcortical surfaces of ADNI participants.

>>>>

With this background in place, we turn to the motivating situation of analyzing the subcortical surfaces of ADNI participants using MR images, with a focus on the hippocampus. ADNI is a longitudinal observational study, where several imaging modalities are used to record neuroimages from participants over time. Participants can generally be considered to belong to three groups based on cognitive state: elderly healthy controls, those displaying symptoms of mild cognitive impairment, and those diagnosed with AD. In practice, participants may transition between these categorizations as their symptoms progress. However, because patients may be diagnosed with AD at different stages, accurately modeling these transitions is challenging and is beyond the scope of this article. For simplicity, we use the groupings recorded at the time each participant was recruited to the study.

Individual participants are typically scheduled for imaging at six month intervals, though variability exists in terms of the recorded intervals. Also of note, the numbers of observations contributed by each participant varies, so groups may be unbalanced. We denote the total number of participants as $N$, and use $n_i$ to represent the number of images present for each individual. Meanwhile, $G$ denotes the number of observed groups under consideration.

Within a segmented structural MRI, the voxels estimated to mark the surface of the hippocampus can be assumed to lie along a two-dimensional manifold that has been wrapped into the three-dimensional image space. More generally, we let $d$ denote the intrinsic dimension of the low-dimensional manifold, and $D$ represent the dimension of the high-dimensional input space. For the $j$th image from the $i$th individual belonging to group $g$ with $i = 1, \dots, N$, $j = 1, \dots, n_i$, and $g = 1, \dots, G$, we denote the $k$th voxel in the high-dimensional space as $\mathbf{x}_{ijk} \in \mathbb{R}^{3}$, and its associated low-dimensional parameterization as $\mathbf{y}_{ijk} \in \mathbb{R}^2$. We let $t_{ij}$ denote the time at which the $j$th image for individual $i$ is recorded. The embedding function which maps from the $d$-dimensional manifold to $D$-dimensional ambient space is represented as $\psi(\mathbf{y}_{ijk})$, which can be modeled as \[%
  \mathbf{x}_{ijk} = \psi(\mathbf{y}_{ijk}) = f_p(\mathbf{y}_{ijk}, t_{ij}) + f_{g(i)}(\mathbf{y}_{ijk}, t_{ij}) + f_i(\mathbf{y}_{ijk}, t_{ij}) + f_{ij}(\mathbf{y}_{ijk}) + \epsilon
.\]%

Each function in this additive model represents a different level of nesting in the data, with $f_p(\mathbf{y}, t)$ estimating the embedding at the population level, and $f_g(\mathbf{y}, t)$, $f_i(\mathbf{y}, t)$, and $f_{ij}(\mathbf{y})$ describing deviations from the population estimates specific to the group, individual, and image in question, respectively. As estimating longitudinal changes in the underlying manifold may be of interest, $f_p(\mathbf{y}, t)$, $f_g(\mathbf{y}, t)$ and $f_i(\mathbf{y}, t)$ are each denoted as varying-coefficient models (\cite{hastieVaryingCoefficientModels1993}). Thus, the model is capable of estimating separate longitudinal changes for population, group, and individual estimates. In cases where longitudinal changes are not of interest, each varying-coefficient model may be constrained to be held constant with respect to time. As the $ij$th image is recorded at a single time point, $f_{ij}(\mathbf{y}, t)$ may not be dependent on time. We specify $f_{ij}(\mathbf{y})$ as a smoothing spline, while the functions $f_p(\mathbf{y}, t)$, $f_g(\mathbf{y}, t)$, and $f_i(\mathbf{y}, t)$ are each varying-coefficient smoothing spline models, estimated by using a smoothing spline to smooth over smoothing spline coefficients calculated at each time point available in the data.

As mentioned previously, this proposed additive spline model structure has several benefits.
